From the function-sheaf dictionary to quasicharacters of $p$-adic tori

### 2015-07-21 dictionary.tex
Modifications to the discrete isogenies section.

A finite, \'etale, surjective morphism $H\to G$ of smooth group schemes over $\Fq$ for which the action of $\Gal(\bFq/\Fq)$ on the kernel is trivial is called a {\it discrete isogeny}, following \cite{kamgarpour:09a}*{\S 2.2}.

@@

We remark that, since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.

@@

\begin{proof}
Objects in $\bCS(G)$ may be described by a small modification to the technique used in Sections~\ref{ssec:E} and \ref{ssec:S}. 
Set $F^{i,j} \ceq C^i_{\text{cts}} (\Gal(\bFq/\Fq), C^j(G(\bFq), \EEx))$.
Then the results of Section~\ref{ssec:E} adapt to give a short exact sequence in continuous Galois cohomology
 \[
    0 \to
    \Hh^0(\Fq,\Hh^2(\bG,\EEx)) \to
    \Hh^2(F^\bullet_G) \to
    \Hh^1(\Fq,\Hh^1(\bG,\EEx)) \to
    0,
 \]
for which the maps are given by the analogues of Proposition~\ref{prop:ses_desc}.
Moreover, using \cite{deligne-katz:SGA7.2}*{Expos\'e XIII, Rappel 1.1.3} we see that Proposition~\ref{prop:SGiso} adapts to provide an isomorphism $\bCSiso{G}\to \Hh^2(F^\bullet_G)$ compatible with $\bCS(G) \to \CS(G)$ and with the
the canonical map of exact sequences 
 \[
\begin{tikzcd}
    0 \arrow{r} &
    \Hh^0(\Fq,\Hh^2(\bG,\EEx)) \arrow{r} \arrow{d} &
    \Hh^2(F^\bullet_G) \arrow{r} \arrow{d} &
    \Hh^1(\Fq,\Hh^1(\bG,\EEx)) \arrow{r} \arrow{d} &
    0
\\
    0 \arrow{r} & 
    \Hh^0(\Weil{},\Hh^2(\bG,\EEx)) \arrow{r} & 
    \Hh^2(E^\bullet_G) \arrow{r} & 
    \Hh^1(\Weil{},\Hh^1(\bG,\EEx)) \arrow{r} &
    0.
\end{tikzcd}
 \]
In this way, Proposition~\ref{prop:bounded-etale} is now reduced to the claim
\[
\Hh^1(\Fq,\Hh^1(\bG,\EEx)) = \Hom(G(\Fq),\ZEx).
\]
To see that, one may argue as follows. 
Pick $i\in \pi_0(G)$ and let $G^i \hookrightarrow G$ be the corresponding connected component. 
Pick a geometric point ${\bar x}^i$ on $G^i$ and observe that since $G^i$ is connected as a $\Fq$-scheme, $G^i(\bFq)$ is canonically identified with the $\Gal(\bFq/\Fq)$-orbit of ${\bar x}^i$. 
We remark that while $G^i$ is defined over $\Fq$, the set $G^i(\Fq)$ is non-empty only when $G^i(\bFq) = \{ {\bar x}^i\}$.
Since $\Hh^1(\bG,\EEx) = \Hom(\bG,\EEx)$, evaluation $\chi \mapsto \chi({\bar x}^i)$ defines $\Hh^1(\Fq,\Hh^1(\bG,\EEx)) \to \Hh^1(\Fq,\EEx)$. 
By continuity, $\Hh^1(\Fq,\EEx) = \Hh^1(\Fq,\ZEx)$.
Letting $i$ range over $\pi_0(G)$ we conclude that $\Hh^1(\Fq,\Hh^1(\bG,\EEx)) = \Hh^1(\Fq,\Hh^1(\bG,\ZEx))$.
When adapted to abelian groups with continuous action of $\Gal(\bFq/\Fq)$, the strategy of the proof of Lemma~\ref{lem:dual-inv} gives $\Hh^1(\Fq,\Hh^1(\bG,\ZEx)) =  \Hom(G(\Fq),\ZEx)$, concluding the proof.
\end{proof}

This simple observation reveals the necessity of working with Weil sheaves in Definition~\ref{def:CS}: one cannot geometrize all characters of $G(\Fq)$ using local systems on $G$, for general smooth commutative groups schemes $G$.
Proposition~\ref{prop:bounded-etale} is extended to all smooth commutative groups schemes in Section~\ref{ssec:revisited}.

@@

While most of this paper focuses on the question of determining the isomorphism classes of objects, a complete understanding of the morphisms in $\CS(G)$ also requires a description of the automorphisms of an arbitrary character sheaf $\cs{L}$.

@@

We now wish to apply the results of Section~\ref{ssec:connected} to the identity component of $G$; to do so, we must know that the identity component of $G$ is actually an algebraic group over $\Fq$.

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth commutative group scheme over $\Fq$ then
  $\TrFrob{G} : \CSiso{G} \to G(\Fq)^*$ is surjective, admits a section, and has kernel canonically isomorphic to $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$:
    \[
  \begin{tikzcd}
0 \arrow{r} & \Hh^2(\pi_0(\bG),\EEx)^{\Weil{}} \arrow{r} & \CSiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0. 
  \end{tikzcd}
\]
is a split exact sequence.
\end{theorem}

@@

\begin{remark}
Although $\TrFrob{\pi_0(G)}$ is split and $\TrFrob{G^0}$ is an isomorphism, it does not follow that $\TrFrob{G}$ is split. 
\end{remark}

@@

\begin{proposition} \label{prop:bounded}
Let $G$ be a smooth commutative group scheme over $\Fq$. 
Then $\cs{L}\in \CS(G)$ descends to $G$ if and only if $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ has bounded image.
\end{proposition}
\begin{proof} 
By Lemma~\ref{lem:G0alg-grp}, the identity component $G^0$ is a connected algebraic group over $\Fq$. 
It follows from Proposition~\ref{prop:connected} that the restriction of $\cs{L}$ to $G^0$ descends to $G$. 
Also, since $G^0(\Fq)$ is finite, the image of $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ is a finite subgroup and therefore bounded image.  
%
If $\chi \in G(\Fq)^*$ then there is some finite-image character $\chi_0$
with the same restriction to $G^0(\Fq)$ since $G^0(\Fq)$ is lies inside the torsion part of
the finitely generated abelian group $G(\Fq)$.  Therefore $\chi$ is bounded 
if and only if $\chi \cdot \chi_0^{-1}$ is bounded.  But $\chi \cdot \chi_0^{-1}$ descends
to a character of $\pi_0(G)$.
Thus, it is enough to prove Corollary~\ref{prop:bounded}
for \'etale group schemes $G$, which is done in Proposition~\ref{prop:bounded-etale}.
\end{proof}

Proposition~\ref{prop:bounded} shows that the full subcategory
$\bCS(G) \subset \CS(G)$ is not an equivalence, for smooth commutative group schemes $G$.

@@

Having extended the function--sheaf dictionary from connected commutative algebraic groups over $\Fq$ to smooth commutative group schemes $G$ over $\Fq$, we briefly look back to the case when $G$ is a commutative algebraic group. 
Although Weil sheaves are not necessary in that case, the dictionary is still not perfect, generally.

>>>>

A finite, \'etale, surjective morphism $H\to G$ of smooth group schemes over $\Fq$ for which the action of $\Gal(\bFq/\Fq)$ on the kernel is trivial is called a {\it discrete isogeny}, inspired by \cite{kamgarpour:09a}*{\S 2.2}.

@@

\begin{remark}
Since $V$ is $1$-dimensional, the choice of $V \to V\otimes V$ is exactly the choice of an isomorphism $V\iso \EE$.
\end{remark}

\begin{remark}
A descent argument similar to the one employed in the proof of Lemma~\ref{prop:finite} is used in \cite{Drinfeld:math/0609769}*{Lemma~1.10}, though in the more restrictive case of connected algebraic groups.
\end{remark}

@@

\begin{proof}
Objects in $\bCS(G)$ may be described by a small modification to the technique used in Sections~\ref{ssec:E} and \ref{ssec:S}. 
Set $F^{i,j} \ceq C^i_{\text{cts}} (\Gal(\bFq/\Fq), C^j(G(\bFq), \EEx))$.
Then the results of Section~\ref{ssec:E} adapt to give a short exact sequence in continuous Galois cohomology
 \[
    0 \to
    \Hh^0(\Fq,\Hh^2(\bG,\EEx)) \to
    \Hh^2(F^\bullet_G) \to
    \Hh^1(\Fq,\Hh^1(\bG,\EEx)) \to
    0,
 \]
for which the maps are given by the analogues of Proposition~\ref{prop:ses_desc}.
Moreover, using \cite{deligne-katz:SGA7.2}*{Expos\'e XIII, Rappel 1.1.3} we see that Proposition~\ref{prop:SGiso} adapts to provide an isomorphism $\bCSiso{G}\to \Hh^2(F^\bullet_G)$ compatible with $\bCS(G) \to \CS(G)$ and with the
the canonical map of exact sequences 
 \[
\begin{tikzcd}
    0 \arrow{r} &
    \Hh^0(\Fq,\Hh^2(\bG,\EEx)) \arrow{r} \arrow{d} &
    \Hh^2(F^\bullet_G) \arrow{r} \arrow{d} &
    \Hh^1(\Fq,\Hh^1(\bG,\EEx)) \arrow{r} \arrow{d} &
    0
\\
    0 \arrow{r} & 
    \Hh^0(\Weil{},\Hh^2(\bG,\EEx)) \arrow{r} & 
    \Hh^2(E^\bullet_G) \arrow{r} & 
    \Hh^1(\Weil{},\Hh^1(\bG,\EEx)) \arrow{r} &
    0.
\end{tikzcd}
 \]
In this way, Proposition~\ref{prop:bounded-etale} is now reduced to the claim
\[
\Hh^1(\Fq,\Hh^1(\bG,\EEx)) = \Hom(G(\Fq),\ZEx).
\]
To see that, one may argue as follows. 
Pick $i\in \pi_0(G)$ and let $G^i \hookrightarrow G$ be the corresponding connected component. 
Pick a geometric point $x$ on $G^i$ and observe that since $G^i$ is connected as a $\Fq$-scheme, $G^i(\bFq)$ is canonically identified with the $\Gal(\bFq/\Fq)$-orbit of $x$. 
We remark that while $G^i$ is defined over $\Fq$, the set $G^i(\Fq)$ is non-empty only when $G^i(\bFq) = \{ x\}$.
Since $\Hh^1(\bG,\EEx) = \Hom(\bG,\EEx)$, evaluation $\chi \mapsto \chi(x)$ defines $\Hh^1(\Fq,\Hh^1(\bG,\EEx)) \to \Hh^1(\Fq,\EEx)$. 
By continuity, $\Hh^1(\Fq,\EEx) = \Hh^1(\Fq,\ZEx)$.
Letting $i$ range over $\pi_0(G)$ we conclude that $\Hh^1(\Fq,\Hh^1(\bG,\EEx)) = \Hh^1(\Fq,\Hh^1(\bG,\ZEx))$.
When adapted to abelian groups with continuous action of $\Gal(\bFq/\Fq)$, the strategy of the proof of Lemma~\ref{lem:dual-inv} gives $\Hh^1(\Fq,\Hh^1(\bG,\ZEx)) =  \Hom(G(\Fq),\ZEx)$, concluding the proof.
\end{proof}

Proposition~\ref{prop:bounded-etale} reveals the necessity of working with Weil sheaves in Definition~\ref{def:CS}: one cannot geometrize all characters of $G(\Fq)$ using local systems on $G$, for general smooth commutative groups schemes $G$.
Proposition~\ref{prop:bounded-etale} is extended to all smooth commutative groups schemes in Section~\ref{ssec:revisited}.

@@

%While most of this paper focuses on the question of determining the isomorphism classes of objects, 
A complete understanding of the morphisms in $\CS(G)$ also requires a description of the automorphisms of an arbitrary character sheaf $\cs{L}$.

@@

We now wish to apply the results of Section~\ref{ssec:connected} to the identity component of $G$, for which we must confirm that the identity component of $G$ is actually an algebraic group over $\Fq$.

@@

\begin{theorem}\label{thm:snake}
  If $G$ is a smooth commutative group scheme over $\Fq$ then
  $\TrFrob{G} : \CSiso{G} \to G(\Fq)^*$ is surjective and has kernel canonically isomorphic to $\Hh^2(\pi_0(\bG),\EEx)^{\Weil{}}$:
    \[
  \begin{tikzcd}
0 \arrow{r} & \Hh^2(\pi_0(\bG),\EEx)^{\Weil{}} \arrow{r} & \CSiso{G} \arrow{r}{\TrFrob{G}} & G(\Fq)^* \arrow{r} & 0. 
  \end{tikzcd}
\]
is an exact sequence.
\end{theorem}

@@

\begin{remark}\label{rem:rats}
Although $\TrFrob{\pi_0(G)}$ is split and $\TrFrob{G^0}$ is an isomorphism, we do not know if $\TrFrob{G}$ is split, in general. Surjectively of $\TrFrob{G}$ shows that every $\ell$-adic character of $G(\Fq)$ admits a geometrization, but without a splitting for $\TrFrob{G}$ we do not know how to make this geometrization canonical.
\end{remark}

@@

\begin{proposition} \label{prop:bounded}
Let $G$ be a smooth commutative group scheme over $\Fq$. 
Then $\cs{L}\in \CS(G)$ descends to $G$ if and only if $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ has bounded image.
\end{proposition}
\begin{proof} 
By Lemma~\ref{lem:G0alg-grp}, the identity component $G^0$ is a connected algebraic group over $\Fq$. 
It follows from Proposition~\ref{prop:connected} that the restriction of $\cs{L}$ to $G^0$ descends to $G$. 
Also, since $G^0(\Fq)$ is finite, the image of $\trFrob{\cs{L}} : G(\Fq) \to \EEx$ is a finite subgroup and therefore has bounded image.  
%
If $\chi \in G(\Fq)^*$ then there is some finite-image character $\chi_0$
with the same restriction to $G^0(\Fq)$ since $G^0(\Fq)$ is lies inside the torsion part of
the finitely generated abelian group $G(\Fq)$.  Therefore $\chi$ is bounded 
if and only if $\chi \cdot \chi_0^{-1}$ is bounded.  But $\chi \cdot \chi_0^{-1}$ descends
to a character of $\pi_0(G)$.
Thus, it is enough to prove Corollary~\ref{prop:bounded}
for \'etale group schemes $G$, which is done in Proposition~\ref{prop:bounded-etale}.
\end{proof}

Proposition~\ref{prop:bounded} shows that the full subcategory
$\bCS(G) \subset \CS(G)$ is not an equivalence, for general smooth commutative group schemes $G$. Again we see the necessity of working with Weil sheaves in Definition~\ref{def:CS}.

@@

Having extended the function--sheaf dictionary from connected commutative algebraic groups over $\Fq$ to smooth commutative group schemes $G$ over $\Fq$, we look back briefly to the case when $G$ is a commutative algebraic group. 
Although Weil sheaves are not necessary in that case, the dictionary is still not perfect, generally.
